---
title: "Procrastination & Dementia"
subtitle: "Using Markov models to predict transitions to dementia"
---

```{r}
#| warning: false
#| message: false
#| label: set-up

# Packages ---------------------------------------------------------------------
library(dplyr)
library(tidyr)
library(ggplot2)

# Functions --------------------------------------------------------------------
source(here::here("R/Functions.R"))

# Attributes -------------------------------------------------------------------
colour = "#212427"

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = rel(2), colour = colour, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = rel(1.75), colour = colour),
      axis.title = element_text(size = rel(1.5), colour = colour, face = "bold"),
      strip.text = element_text(size = rel(1), colour = colour, face = "bold"),
      legend.title = element_text(hjust = 0.5, colour = colour, face = "bold"),
      ))

# Data -------------------------------------------------------------------------
df <- read.csv(here::here("data/data.csv"))

# Covariates -------------------------------------------------------------------
covariates <- c(
  "ID", "Gender", "Age", "Education_tri", "Total_dep_2016", "Total_p")
```

## Plan for today

-   Linking procrastination to dementia
-   Markov models
    -   Using Markov models for dementia
    -   Data analysis

::: fragment
-   **Time permitting: Goodness of fit metrics**
:::

# **Dementia & Procrastination** {background-color="#40666e"}

## Dementia & Procrastination

### What is dementia

::: incremental
-   Dementia is a syndrome characterized by the [progressive]{.alert} and [typically irreversible decline]{.alert} of cognitive function.
-   It encompasses a range of conditions, including Alzheimer’s disease, vascular dementia, and Lewy body dementia.
-   As a major public health concern, dementia impacts [millions worldwide]{.alert} [@Nichols2022], straining healthcare systems and caregivers.
-   Given its growing prevalence, [identifying modifiable risk factors]{.alert} is essential for prevention and early intervention
:::

## Dementia & Procrastination

### The Role of Mild Cognitive Impairment

::: incremental
-   MCI involves cognitive changes—such as [memory lapses or impaired decision-making]{.alert}—that exceed normal aging but don’t yet meet dementia criteria.
-   Often a [precursor to dementia]{.alert} [@Cooper2015; @Shigemizu2020; @Tschanz2006], MCI significantly raises the risk of progression, making early detection and intervention critical
:::

## Dementia & Procrastination

### Apathy: A Key Behavioral Symptom

::: incremental
-   Among the most common behavioral changes in MCI and dementia is apathy—a loss of motivation and goal-directed behavior.
-   Research shows apathy [nearly doubles the risk of progressing to dementia]{.alert} [@vanDalen2018], highlighting its role as a potential early marker
:::

## Dementia & Procrastination

### Procrastination: A Parallel Dysfunction?

::: incremental
-   Unlike apathy, procrastination involves delaying tasks despite intent, suggesting a conflict in executive function rather than pure disengagement.
-   Both behaviors point to impairments in goal-oriented decision-making, a [core deficit in MCI and dementia]{.alert}.
:::

:::: fragment
::: callout-tip
## Research question

Could procrastination, like apathy, signal broader cognitive and motivational decline?
:::
::::

# **Markov Models** {background-color="#40666e"}

## Markov Models

### Core Concept

To analyze longitudinal patterns of cognitive decline, we used a [discrete-time, first-order Markov model]{.alert}.

::: incremental
-   The model hinges on the [Markov property]{.alert}:
    -   Let $X_t$ denote an individual’s cognitive state at time $t$, where $X_t \in S = \{1, 2, \dots, K\}$.
    -   Transition to state $j$ at $t + 1$ depends [only]{.alert} on state $i$ at $t$.
:::

:::: fragment
::: callout-note
## Markov property

$$P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)$$
:::
::::

## Markov Models

### Transition Matrices

These probabilities, denoted $p_{ij}$ are called [transition probabilities]{.alert}

::: panel-tabset
### Stationary Model

$$
P = \begin{bmatrix} 
p_{11} & \cdots & p_{1K} \\
\vdots & \ddots & \vdots \\
p_{K1} & \cdots & p_{KK} \\
\end{bmatrix}
$$

### Non-Stationary Model

$$
P^{(t)} = \begin{bmatrix} 
p_{11}^{(t)} & \cdots& p_{1K}^{(t)} \\
\vdots & \ddots & \vdots \\
p_{K1}^{(t)} & \cdots & p_{KK}^{(t)} \\
\end{bmatrix}
$$

### Our Model

$$
P^{(t)} = \begin{bmatrix} 
p_{11}^{(t)} & p_{12}^{(t)} & p_{13}^{(t)} \\
p_{21}^{(t)} & p_{22}^{(t)} & p_{23}^{(t)} \\
0      & 0      & 1      \\
\end{bmatrix}
$$
:::

-   **Rows**: Current state $(t)$
-   **Columns**: Next state $(t + 1)$

::: footer
**Formally:** $p_{ij} = P(X_{t+1} = j \vert X_t = i)$
:::

## Markov Models

### Discrete Time Models

-   We capture time-varying transition probabilities using [multinomial logistic regression]{.alert}

::: fragment
1.  **Model Specification** (for $K=3$ states with state $K$ as reference):
    -   Log-odds of transitioning to state $j$ vs. state $K$:
:::

::: fragment
$$
\log \left( \frac{p_{ij}^{(t)}}{p_{iK}^{(t)}} \right) = \underbrace{\alpha_j^{(t)}}_{\text{intercept}} + \underbrace{\boldsymbol{\beta}_j^{(t)^T} \boldsymbol{X}^{(t)}}_{\text{covariate effects}} \quad \text{for } j = 1, \dots, K-1
$$
:::

# **Data Analysis** {background-color="#40666e"}

## Data Analysis

### Our Dataset

-   Analyses were conducted using a secondary data source called the [Health and Retirement Study (HRS)]{.alert}
-   We focused on four biennial waves of HRS data (2016 - 2022).

::: fragment
```{r}
#| label: tbl-cog-data
#| tbl-cap: "Cognition data from the HRS (2016 - 2022)"
#| tbl-cap-location: bottom

data <- df |>
  extract_years(years = seq(2016, 2022, by = 2), impute = TRUE, absorbing = TRUE) |>
  na.omit()

head(data) |>
  rename_with(~ stringr::str_replace(., "cogfunction", "")) |>
  knitr::kable(format = "html")
```
:::

## Data Analysis

### Our Dataset

```{r}
#| fig-width: 20
#| fig-height: 10
#| label: fig-dementia-transitions
#| fig-cap: "Cognitive transitions across time (with dementia as an absorbing state)"

df |>
  count_transitions(years = seq(2016, 2022, by = 2), absorbing = TRUE) |>
  group_by(ID, Wave) |>
  reframe(plyr::count(Classification)) |>
  rename(Classification = x, n = freq) |>
  plot_transitions(size = rel(4.5)) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(size = rel(1.5)),
    legend.position = "none"
  )
```

## Data Analysis

### Our Dataset

```{r}
#| echo: true
#| output-location: column
#| code-line-numbers: "1-7"
#| label: tbl-pivot-data
#| tbl-cap: "Pivoted data"
#| tbl-cap-location: bottom

# Pivoting to long format
data_long <- data |>
  dplyr::inner_join(df[, covariates], by = "ID") |>
  pivot_and_factorise() |>
  dplyr::group_by(ID) |>
  dplyr::mutate(Age = Age - (2022 - as.numeric(as.character(wave)))) |>
  dplyr::ungroup()

head(data_long, n = 8) |>
  dplyr::select(ID, wave, status, Total_p) |>
  knitr::kable(
    format = "html", align = "c",
    col.names = c(
      "ID", "Wave", 
      "Status", "Procrastination")) |>
  kableExtra::kable_paper(
    html_font = "Arial", 
    font_size = 20)
```

## Data Analysis

### Our Dataset

```{r}
#| echo: true
#| output-location: column
#| code-line-numbers: "3"
#| label: tbl-stacked-data
#| tbl-cap: "Data with new t - 1 column"
#| tbl-cap-location: bottom

data_stack <- data_long |>
  dplyr::group_by(ID) |>
  dplyr::mutate(status_prev = dplyr::lag(status), .after = status) |>
  dplyr::filter(!is.na(Total_p), Age >= 50) |>
  dplyr::ungroup() |>
  dplyr::filter(wave != 2016)

head(data_stack, n = 8) |>
  dplyr::select(ID, wave, status, status_prev, Total_p) |>
  knitr::kable(
    format = "html", align = "c",
    col.names = c(
      "ID", "Wave", 
      "Status (t)", "Status (t-1)", 
      "Procrastination")) |>
  kableExtra::kable_paper(
    html_font = "Arial", 
    font_size = 20)
```

## Data Analysis

### Transition Frequencies

```{r}
#| fig-width: 12
#| label: fig-frequencies
#| fig-cap: "Transition frequencies between cognitive states for each time period."

# Creating a table dataset
table_data <- df |>
  extract_years(seq(2016, 2022, by = 2), impute = TRUE, absorbing = TRUE) |>
  rename_with(~ gsub("cogfunction", "HRS_", .)) |>
  mutate(
    across(c(HRS_2016:HRS_2022), ~ factor(.x, levels = c("Normal Cognition", "MCI", "Dementia")))
  )

# Creating transition frequencies ---------------------------------------------
## These are the time periods we are interested in
time_periods <- list(
  c("2016", "2018"),
  c("2018", "2020"), 
  c("2020", "2022")
)

# Applying function
transition_results <- purrr::map(time_periods, ~ create_transition_table(.x[1], .x[2]))
names(transition_results) <- purrr::map_chr(time_periods, ~ paste(.x[2], .x[1], sep = "-"))

### Creating one dataset
transition_frequencies <- time_periods |>
  create_transition_dataset(transition_results = transition_results)

transition_frequencies |> 
  tran_heat_map() +
  theme(plot.title = element_text(size = 12),
        axis.title = element_text(size = 10))
```

## Data Analysis

### Modelling

::::: panel-tabset
### Stationary

```{r}
#| echo: true
#| code-line-numbers: "1-4|6-9|11-14"

# Model 1: Baseline model (reference: NC)
fit_1a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016,
  family = multinomial, data = data_stack, trace = FALSE)

# Model 2: With procrastination (reference: NC)
fit_2a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p,
  family = multinomial, data = data_stack, trace = FALSE)

# Model 3: Full model with previous state (reference NC)
fit_3a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, 
  family = multinomial, data = data_stack, trace = FALSE)
```

```{r}
fit_3b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, 
  family = multinomial, data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)
```

### Non Stationary

```{r}
#| echo: true
#| code-line-numbers: "1-4|6-9|11-14"

# Model 4: Additive time effects
fit_4a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev + wave, 
  family = multinomial, data = data_stack, trace = FALSE)

# Model 5: State-specific time effects
fit_5a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + (status_prev * wave), 
  family = multinomial, data = data_stack, trace = FALSE)

# Model 6: Full time interactions
fit_6a <- nnet::multinom(
  status ~ (Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev) * wave, 
  family = multinomial, data = data_stack, trace = FALSE)
```

### Comparison

::: {style="font-size: 75%; text-align: center;"}
@tbl-stationary-comparison & @tbl-non-stationary-comparison present [likelihood ratio tests]{.alert} for each model
:::

```{r}
#| label: tbl-stationary-comparison
#| tbl-cap: "Likelihood ratio test for stationary models"
#| tbl-cap-location: bottom

anova(fit_1a, fit_2a, fit_3a) |>
  mutate(
    Model = c("Baseline", "Procrastination", "Procrastination + Previous Status"),
    `Resid. Dev`= round(`Resid. Dev`, digits = 3),
    `LR stat.` = round(`LR stat.`, digits = 3),
    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
    ) |>
  knitr::kable(
    format = "html", align = "c") |>
  kableExtra::kable_paper(
    html_font = "Arial", 
    font_size = 20)
```

::: fragment
```{r}
#| label: tbl-non-stationary-comparison
#| tbl-cap: "Likelihood ratio test for non-stationary models"
#| tbl-cap-location: bottom

anova(fit_3a, fit_4a, fit_5a, fit_6a) |>
  mutate(
    Model = c("Stationary", "Additive", "State-Time Interactions", "Full Interactions"),
    `Resid. Dev` = round(`Resid. Dev`, digits = 3),
    `LR stat.` = round(`LR stat.`, digits = 3),
    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
    ) |>
    knitr::kable(
    format = "html", align = "c") |>
  kableExtra::kable_paper(
    html_font = "Arial", 
    font_size = 20)
```
:::
:::::

::: footer
$\text{LRT} = -2 \times \ell_{\text{reduced}} - \ell_{\text{full}}$
:::

## Data Analysis

### Visualisation

::: panel-tabset
### Odds Ratios

```{r}
#| fig-width: 12
#| label: fig-odds-ratio
#| fig-cap: "Estimated odds ratios for the stationary model."

stationary_results <- rbind(tidy_output(fit_3a), tidy_output(fit_3b)) |>
  rename(transition = y.level) |>
  mutate(
    # Transforming to odds ratios
    estimate = exp(estimate),
    conf.low = exp(conf.low),
    conf.high = exp(conf.high),
    
    # Mutating to factors
    transition = factor(
      transition, 
      levels = c("NC - MCI", "MCI - NC", 
                 "NC - Dementia", "MCI - Dementia")),
    term = factor(
      term, 
      levels = c("Being female", "Age", "High school degree vs. No education",
                 "Further education vs. No education", "Depression Scores (2016)",
                 "Procrastination (2020)", "Previous state: MCI", 
                 "Previous state: Dementia")),
    
    # Creating a colour code
    colour = case_when(
      estimate > 1 & p.value < 0.05 ~ "Positive",
      estimate < 1 & p.value < 0.05 ~ "Negative",
      TRUE ~ "NS"),
    
    colour = factor(colour, levels = c("Positive", "Negative", "NS"))
    )

fig_4 <- stationary_results |>
  filter(!term %in% c("Previous state: MCI", "Previous state: Dementia")) |>
  ggplot(aes(x = estimate, y = transition, colour = colour)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  ggstance::geom_pointrangeh(
    aes(xmin = conf.low, xmax = conf.high),
    position = ggstance::position_dodgev(height = 0.5),
    size = 1.25,
    fatten = 3) +
  scale_colour_manual(values = c(
    "Positive" = "#0072B2", 
    "Negative" = "#E69F00", 
    "NS"       = "#B2BEB5")) +
  labs(title = "Odds ratios - Stationary model",
       x = "Odds Ratio", y = "Transition") +
  guides(colour = "none") +
  facet_wrap(~ term, scales = "free_x") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank())

fig_4
```

### Predictions

```{r}
#| fig-width: 12
#| label: fig-model-predictions
#| fig-cap: "Predicted transition probabilities from the stationary model."

pred_data <- expand.grid(
  Gender = factor(0),
  Age = mean(data_stack$Age),
  Education_tri = factor(0),
  Total_dep_2016 = mean(data_stack$Total_dep_2016),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200))

## Plotting predictions
fig_5 <- pred_data |>
  modelr::add_predictions(model = fit_3a, var = "pred", type = "probs") |>
  tidy_predictions() |>
  plot_predictions_stationary(variable = Total_p, x_axis = "Total Procrastination") +
  theme(plot.title = element_text(size = 14),
        axis.title = element_text(size = 12))

fig_5
```

### Estimated Matrix

```{r}
#| fig-width: 12
#| label: fig-matrix-comparison
#| fig-cap: "Estimated transition matrix."

# Get all unique states
names <- c("Normal Cognition", "MCI", "Dementia")
states <- sort(unique(data_stack$status))
n_states <- length(states)

# Create empty transition matrix
transition_matrix_estimated <- matrix(
  0, nrow = n_states, ncol = n_states,
  dimnames = list(paste("From", names), paste("To", names))
  )

# Getting estimated probabilities
estimated_probs <- expand.grid(
  Gender = factor(0),
  Age = mean(data_stack$Age),
  Education_tri = factor(0),
  Total_dep_2016 = mean(data_stack$Total_dep_2016),
  Total_p = mean(data_stack$Total_p),
  status_prev = states) |>
  modelr::add_predictions(model = fit_3a, var = "pred", type = "probs")

# Filling in matrix
for(i in 1:n_states) {
  transition_matrix_estimated[i, ] <- estimated_probs$pred[i, ]
}

fig_6 <- transition_matrix_estimated |>
  reshape_matrix() |>
  plot_transition_matrix(observed = FALSE) +
  theme(plot.title = element_text(size = 14),
        plot.subtitle = element_text(size = 12),
        axis.title = element_text(size = 12))

fig_6
```
:::

# **Conclusion** {background-color="#40666e"}

## Conclusion
### Wrapping up

-   Preliminary results suggest that procrastination is associated with cognitive decline.
-   Older adults with higher levels of procrastination have:
    -   [An increased probability of transitioning to MCI.]{.alert}
    -   [A decreased probability of transitioning back to NC]{.alert}
-   Procrastination is a complex behavior that may signal [broader cognitive and motivational decline]{.alert}.

## References

::: {#refs}
:::

## Wrapping up
### Thank you

::: {style="text-align:center; font-size:1.5em;"}
![These slides were built with <br> [{{< fa brands github title="GitHub octocat logo" >}}](https://c-monaghan.github.io/), [{{< fa brands r-project title="R Project" >}}](https://www.r-project.org/about.html){target="_blank"}, and [Quarto](https://quarto.org/){target="_blank"}](./files/images/QR_Code.svg){fig-align="center" width="450" height="450"}
:::
