---
title: "Procrastination as a Behavioural Marker of Cognitive Decline"
subtitle: "Evidence from Longitudinal Transitions in Older Adults"
---


```{r}
#| label: setup
#| warning: false
#| message: false

# Packages ---------------------------------------------------------------------
pacman::p_load(
  dplyr,
  tidyr,
  ggplot2,
  ggalluvial,
  patchwork,
  catpuccin
)

# Functions --------------------------------------------------------------------
functions <- list.files(path = "R/", pattern = "\\.R$", full.names = TRUE)

purrr::walk(functions, source)

# Theme ------------------------------------------------------------------------
# Attributes
colour = "#cdd6f4"

theme_set(
  theme_minimal() +
    theme(
      plot.background = element_rect(fill = "#1e1e2e"),
      panel.grid = element_blank(),
      plot.title = element_text(
        hjust = 0.5,
        size = rel(2.5),
        colour = colour,
        face = "bold"
      ),
      plot.subtitle = element_text(
        hjust = 0.5,
        size = rel(1.75),
        colour = colour
      ),
      axis.title = element_text(
        size = rel(2.5),
        colour = colour,
        face = "bold"
      ),
      axis.text = element_text(
        size = rel(2),
        colour = colour
      ),
      strip.text = element_text(
        size = rel(1),
        colour = colour,
        face = "bold"
      ),
      legend.title = element_text(
        hjust = 0.5,
        colour = colour,
        face = "bold"
      ),
    )
)

# Data -------------------------------------------------------------------------
data <- readRDS(here::here("data/data.RDS")) |> 
  tibble::as_tibble() |>
  rename(Depression = Apathy)

# Making a t-1 column
data_stack <- data |>
  group_by(ID) |>
  mutate(status_prev = lag(status), .after = status) |>
  filter(wave != 2016)
```

# **Dementia & Procrastination**

## Dementia & Procrastination

### What is dementia

::: incremental
-   Syndrome of [progressive, usually irreversible]{.alert} cognitive decline
-   Includes Alzheimer’s, vascular, Lewy body types
-   Affects millions worldwide[@Nichols2022]
-   Focus on [modifiable risks]{.alert} for prevention
:::

::: notes
-   Given that there is currently no cure for dementia, identifying modifiable risk factors is essential for prevention and early intervention.
:::

## Dementia & Procrastination

### The Role of Mild Cognitive Impairment

::: incremental
-   Cognitive changes beyond normal aging - memory lapses, poor decisions
-   Often [precedes]{.alert} dementia[@Cooper2015; @Shigemizu2020; @Tschanz2006]
-   [Early detection crucial]{.alert}
:::

::: notes
**Signs of MCI:**

-   Forgetting things more often.
-   Missing appointments or social events.
-   Finding it hard to make decisions, finish a task or follow instructions.
-   Trouble finding their way around places they know well.
:::

## Dementia & Procrastination

### Apathy vs. Procrastination

::: incremental
-   [Apathy]{.alert}
    -   Loss of motivation
    -   Nearly doubles dementia risk[@vanDalen2018]
-   [Procrastination]{.alert}
    -   Intentional delay despite intent
    -   Signals executive function breakdown
:::

:::: {.fragment .fade-up}
::: callout-tip
## Research question

**Could procrastination signal broader cognitive and motivational decline?**
:::
::::

::: notes
Both behaviours reflect impaired goal-directed behavior
:::

# **Markov Models**

## Markov Models

### Core Concept

To analyze longitudinal patterns of cognitive decline, we used a [discrete-time, first-order Markov model]{.alert}.

::: incremental
-   The model hinges on the [Markov property]{.alert}:
    -   Let $X_t$ denote an individual’s cognitive state at time $t$, where $X_t \in S = \{1, 2, \dots, K\}$.
    -   Transition to state $j$ at $t + 1$ depends [only]{.alert} on state $i$ at $t$.
:::

:::: {.fragment .fade-up}
::: callout-note
## Markov property

$$P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)$$
:::
::::

::: notes
-   Markov models are **memoryless** and rely on the Markov property.
:::

## Markov Models

### Transition Matrices

These probabilities, denoted $p_{ij}$ are called [transition probabilities]{.alert}.

$$
P = \begin{bmatrix} 
p_{11} & \cdots & p_{1K} \\
\vdots & \ddots & \vdots \\
p_{K1} & \cdots & p_{KK} \\
\end{bmatrix}
$$

::: footer
**Formally:** $\; p_{ij} = P(X_{t+1} = j \vert X_t = i)$
:::

## Markov Models

### Discrete Time Models

-   We capture time-varying transition probabilities using [multinomial logistic regression]{.alert}.

::: fragment
1.  **Model Specification** (for $K=3$ states with state $K$ as reference):
    -   Log-odds of transitioning to state $j$ vs. state $K$:
:::

::: fragment

$$
log \left( \frac{P(Y = j \vert x)}{P(Y = k \vert x)} \right) = \underbrace{\alpha}_{\text{intercept}} + \underbrace{\beta_j^T \:X}_{\text{covariate effects}} \qquad \text{for } j = 1, \dots K-1
$$
:::

::: notes
-   $X^{(t)}$ can contain both time variant and time invariant covariates.
:::

# **Data Analysis**

## Data Analysis

### Our Dataset

```{r}
#| label: fig-dementia-transitions
#| fig-height: 15
#| fig-width: 25
#| fig-cap: "Cognitive transitions across time"

data |>
  dplyr::mutate(
    status = dplyr::case_when(
    status == "1" ~ "Normative Cognitive Function", 
    status == "2" ~ "MCI",
    status == "3" ~ "Dementia"),
    status = factor(
      status, levels = c("Normative Cognitive Function", "MCI", "Dementia"))
    ) |>
  dplyr::group_by(ID, wave) |>
  dplyr::reframe(plyr::count(status)) |>
  dplyr::rename(classification = x, n = freq) |>
  ggplot(aes(
    x = wave, y = n, stratum = classification,
    fill = classification, alluvium = ID)) +
  ggalluvial::geom_stratum(alpha = 0.5, width = 0.5) +
  ggalluvial::geom_flow(width = 0.5) +
  geom_text(
    stat = "stratum", aes(label = stringr::str_wrap(classification, width = 10)),
    hjust = 0.5, vjust = 0.5, size = rel(6), colour = colour, fontface = "bold") +
  labs(title = "Cognitive transitions across time", x = "", y = "Frequency") +
  catppuccin::scale_fill_catppuccin(palette = "mocha", discrete = TRUE) +
  theme(legend.position = "none")
```
## Data Analysis

### Modelling

```{r}
#| label: modelling
#| echo: true
#| output-location: fragment
#| code-line-numbers: "1-8"
#| warning: false
#| message: false

fit_a <- nnet::multinom(
  formula = status ~ Gender + Education_tri + Depression + (Age * Total_p) + status_prev, 
  data = data_stack, family = multinomial, trace = FALSE)

fit_b <- nnet::multinom(
  formula = status ~ Gender + Education_tri + Depression + (Age * Total_p) + status_prev, 
  data = data_stack |> mutate(status = relevel(status, ref = "2")), 
  family = multinomial, trace = FALSE)

fit_results <- rbind(tidy_output(fit = fit_a), tidy_output(fit = fit_b))

fit_results |> head() |> output_model()
```
## Data Analysis

### Odds Ratios

```{r}
#| label: fig-model-summary
#| fig-width: 15
#| fig-height: 9
#| fig-cap: "Odds ratios (95% Confidence Intervals) of transitioning from one cognitive state to another"

fit_results |>
  filter(!term %in% c("Previous state: MCI", "Previous state: Dementia")) |>
  mutate(
    label_pos = case_when(
      stringr::str_detect(term, "Further") & 
        stringr::str_detect(transition, "MCI - NC") ~ estimate - 0.5,
      stringr::str_detect(term, "Further") ~ estimate - 0.5,
      TRUE ~ estimate
    ),
    label_text = sprintf(
    "<span style='line-height:3.5'>OR = %.3f<br><br>[%.3f, %.3f]</span>",
    estimate, conf.low, conf.high)) |>
  plot_model() +
  theme(
    axis.title = element_text(
      size = rel(1.5), colour = colour, face = "bold"),
    axis.text = element_text(size = rel(1.5), colour = colour),
    strip.text = element_text(size = rel(1.5), colour = colour),
    panel.spacing = unit(1.5, "line"))
  
```

## Data Analysis

### Predictions

```{r}
#| label: fig-age-predictions
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "Predicted transition probabilities for different age cohorts"

# Create predictions with age info
age_cohots <- expand.grid(
  Gender = factor(0),
  Age = c(62:97),
  Education_tri = factor(0),
  Depression = mean(data_stack$Depression, na.rm = TRUE),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200)) |>
  modelr::add_predictions(model = fit_a, var = "pred", type = "probs") |>
  tidy_predictions() |>
  filter(status_prev %in% c("Normal Cognition", "MCI") &
         status %in%  c("Normal Cognition", "MCI")) |>
  mutate(label = case_when(
    Age %in% c(70, 80, 90) ~ as.character(Age), 
    TRUE ~ "Other"
  ))

# Identifying right most point for each label point
label_data <- age_cohots |>
  filter(Age %in% c(70, 80, 90)) |>
  group_by(status_prev, Age) |>
  filter(Total_p == max(Total_p)) |>
  ungroup()

# Plotting ---------------------------------------------------------------------
fig <- ggplot(data = age_cohots, aes(x = Total_p, y = prob, 
                              group = interaction(Age, status_prev, status),
                              colour = label)) +
  geom_line(aes(linewidth = label)) + 
  ggrepel::geom_text_repel(
    data = label_data,
    aes(label = paste0("Age ", Age), colour = as.character(Age)),
    direction = "y", nudge_x = 4, hjust = 0,
    segment.size = 0.5, segment.colour = "grey50",
    size = 3, show.legend = FALSE) +
  scale_x_continuous(
    breaks = seq(0, 60, by = 10),
    sec.axis = sec_axis( ~., name = "Previous State", breaks = NULL, labels = NULL)) +
  scale_y_continuous(sec.axis = sec_axis( ~ ., name = "Current State", breaks = NULL, labels = NULL)) +
  scale_colour_manual(values = c(
    "70"    = "#cba6f7", 
    "80"    = "#f38ba8", 
    "90"    = "#fab387",
    "Other" = alpha("grey90", 0.2))) +
  scale_linewidth_manual(values = c(
    "70"    = 1.5, 
    "80"    = 1.5, 
    "90"    = 1.5,
    "Other" = 0.5)) +
  labs(
    title = "Predicted transition probabilities",
    x = "Procrastination", y = "Probability", 
    caption = "*Note:* Each curve represents a different age profile, ranging from the minimum (62 years) to the maximum (97 years) of the dataset. <br> Specific ages of interest (70, 80, and 90) are highlighted.",
    colour = NULL) +
  facet_grid(status ~ status_prev) +
  theme(
    plot.caption = ggtext::element_markdown(
      hjust = 0, size = 10, colour = colour),
    axis.title = element_text(size = rel(1.5)),
    axis.text = element_text(size = rel(1.25)),
    strip.text = element_text(size = 10, face = "bold"),
    legend.position = "none",
    panel.grid = element_line(
      colour = alpha(colour, 0.1), size = 0.75, linetype = 2),
    panel.spacing = unit(2.5, "line"))

fig
```


## Conclusions

::: incremental
-   Preliminary evidence that procrastination may function as an [early behavioural marker]{.alert} of cognitive decline, particularly in older age 
-   As a modifiable and measurable construct, procrastination holds promise as a target for [early detection and preventative intervention]{.alert}
:::

## References

::: {#refs}
:::

--- 

<br>

```{r}
#| label: ending
#| fig-width: 15
#| fig-height: 7
#| fig-align: center

fig 
```

